Retrofitting Effect Handlers onto OCaml

### 2021-03-24 retro-concurrency.tex
updates

Effect handlers~\cite{Plotkin09} provide a modular foundation for user-defined
effects. The key idea is to separate the definition of the effectful operations
from their interpretations, which are given by \emph{handlers} of the effects.
For example,
%
\begin{lstlisting}
effect In_line : in_channel -> string
\end{lstlisting}
%
declares an \emph{effect} |In_line|, which is parameterised with an input
channel of type |in_channel|, which when \emph{performed} returns a |string|
value. A computation can perform the |In_line| effect without knowing how the
|In_line| effect is implemented. This computation may be enclosed by different
handlers that handle |In_line| differently. For example, |In_line| may be
implemented by performing a blocking read on the input channel or performing
the read asynchronously by offloading it to an event loop such as |libuv|.
Thanks to the separation of effectful operations from their interpretations,
effect handlers enable new approaches to modular programming.

Effect handlers expose a structured form of delimited
continuations~\cite{Danvy90} that enables non-local control-flow mechanisms
such as resumable exceptions, lightweight threads, coroutines, generators and
asynchronous I/O to be composably expressed. One of the primary motivations to
extend OCaml with effect handlers is to natively support composable
asynchronous I/O to enable highly scalable concurrent applications such as web
servers. Many programming languages, including OCaml, require non-local changes
to source code in order to support asynchronous I/O, often leading to a
dichotomy between synchronous and asynchronous code~\cite{colour}. In the
absence of native support for concurrent programming, OCaml developers use
monadic I/O libraries such as Lwt~\cite{lwt} and Async~\cite{async}, where
asynchronous functions are represented as monadic computations. In these
libraries, while asynchronous functions can call synchronous functions
directly, the converse is not true. In particular, any function that calls an
asynchronous function will also have to be marked as asynchronous. As a result,
large parts of the applications using these libraries tend to be in monadic
form. Languages such as GHC Haskell and Go provide lightweight threads, which
avoids the dichotomy between synchronous and asynchronous code. However, these
languages bake-in the lightweight thread implementation into the runtime
system. With effect handlers, asynchronous I/O can be implemented directly in
OCaml as a library without imposing a monadic form on the users.

@@

OCaml currently does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. A standard way of implementing continuations is
to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Leijen14}. OCaml does not use a CPS IR, and changing
the compiler to utilise a CPS IR would be an enormous undertaking that would
affect the performance profile of existing OCaml applications due to the
increased memory allocations as the continuation closures get allocated on the
heap~\cite{Farvardin20}. Moreover, with CPS, stack unwinding is not meaningful,
and hence, we would lose compatibility with tools that inspect the program
stack. Hence, we choose not to use CPS translation and represent the
continuations as call stacks.

@@

\begin{enumerate}[label=R\arabic*]
	\item \textbf{Backwards compatibility.} A well-behaved OCaml program does
		not break under OCaml extended with effect handlers. OCaml code that does
		not use effect handlers will pay negligible performance and memory cost.

@@

Our work is also timely. The WebAssembly~\cite{Hass17} community group is
considering effect handlers as one of the mechanisms for supporting
concurrency, asynchronous I/O and generators~\cite{WasmProposal}. Project
Loom~\cite{loom} is an OpenJDK project that adds virtual threads and delimited
continuations to Java. The Swift roadmap~\cite{swift} includes direct-style
asynchronous programming and structured concurrency as milestones. We believe
that our design choices will inform similar choices to be made in other
industrial-strength languages.

@@

\begin{itemize}
  \item the design and implementation of one-shot effect handlers for OCaml.
    Our design retains OCaml's compatibility with program analysis tools that
    inspect the stack using DWARF unwind tables. We have validated our DWARF
    unwind tables with the assistance of an automated validator
    tool~\cite{Bastian19}.
	\item a formal operational semantics for the effect handler implementation in
		OCaml. Our formalism explicitly models the subtle interactions with the C
		stack, which is generally overlooked by other formal models, but which the
		implementations must handle.
	\item extensive evaluation which shows that our implementation has negligible
		impact on code that does not use effect handlers, and serves as an
		efficient foundation for scalable concurrent programming.
\end{itemize}

@@

OCaml does not use the C calling convention. In particular, there are no
callee-saved registers in OCaml. In the |x86-64| backend, the OCaml runtime
makes use of two C callee-saved registers for supporting OCaml execution. The
register |r15| holds the \emph{allocation pointer} into the minor heap used for
bump pointer allocation, and |r14| holds a reference to the |Caml_state|, a
table of global variables used by the runtime. This makes external calls
extremely fast in OCaml. If the external function does not allocate in the
OCaml heap, then it can be called directly and no bookkeeping is necessary. For
external functions which allocate in the OCaml heap, the cached allocation
pointer is saved to |Caml_state| before the external call and it is restored on
return. Similarly, callbacks into OCaml are also cheap: these involve loading
the arguments in the right registers and calling the OCaml function. OCaml
callbacks are relatively common as the garbage collector (GC) finalisation
functions are executed as OCaml callbacks.

@@

The lack of callee-saved registers also makes exception handling fast. In the
absence of callee-saved registers, no registers need to be saved when entering
a |try| block. Similarly, no registers need to be restored when handling an
exception. Installing an exception handler simply pushes the program counter
(|pc|) of the handler and the current exception pointer (|exn_ptr| -- a field
in |Caml_state|) to the stack. After this, the current exception pointer is
updated to be the current stack pointer (|rsp|). This creates a linked-list of
exception handler frames on the stack as shown in Figure~\ref{fig:stock_stack}.
Raising an exception simply sets |rsp| to |exn_ptr|, loads the saved |exn_ptr|,
and jumps to the |pc| of the handler.

@@

OCaml generates \emph{stack maps} in order to accurately identify roots on the
stack for assisting the GC. For every call point in the program, the OCaml
compiler emits the size of the frame and the set of all live registers in the
frame containing pointers into the heap. During a GC, the OCaml stack is walked
and the roots are marked, skipping over the C frames.

@@

When the scheduler queue is empty, the |run_next| function performs the pending
reads. We abstract away the details of the event-based I/O using the |do_reads|
function (line 12). |do_reads| takes a list of pending reads and blocks until
at least one of the reads succeeds. It returns a pair of lists |compl| and
|todo|. |compl| contains the result strings from successful reads and
corresponding continuations. These continuations are arranged to be resumed
with the read result and pushed into the scheduler queue. |todo| contains the
channels on which input is still pending and their corresponding continuations.
|pending_reads| is updated to point to the |todo| list so that they may be
attempted later. Observe that all of the changes to add asynchrony are
localised to the |run| function, and the computation that performs these
effects can remain in \emph{direct-style} (as opposed to the monadic-style in
Lwt and Async).

@@

For handling the effects, our implementation extends OCaml's |match ... with|
syntax with effect patterns. The expression
\begin{lstlisting}
match e with
| None -> false | Some b -> b
| effect (E s) k1 -> e1 | effect (F f) k2 -> e2
\end{lstlisting}
\noindent is translated to the equivalent of
\begin{lstlisting}
match_with (fun () -> e)
{ retc = (function None -> false | Some b -> b);
  effc = (function
  | (E s) -> (fun k1 -> e1)
  | (F f) -> (fun k2 -> e2)
  | e -> (fun k -> match perform e with
          | v -> continue k v
          | exception e -> discontinue k e)); }
\end{lstlisting}
For the sake of exposition, we introduce a |('a,'b) handler| type. This handler
handles a |'a comp| that returns a |'a| value, and itself returns a |'b| value.
The handler has a return case |retc| of type |'a -> 'b|. The effect case |effc|
handles effects of type |'c eff| with |('c,'b) continuation| and returns a
value of type |'b|. The last case in |effc| \emph{reperforms} any unmatched
effect to the outer handler and returns the value and exceptions back to the
original performer. In the implementation, reperform is implemented as a
primitive to avoid executing code on the resumption path.

@@

\subsubsection{Syntax} Our expressions (Figure~\ref{sem:syntax}) consist of
integer constants ($n$), variables ($x$), abstraction ($\lam{x}{x}$),
application ($e~e$), arithmetic expressions ($e \odot e$) where $\odot$ ranges
over $\{+,-,*,/\}$, raising exceptions ($\throw{l}{e}$), performing effects
($\perform{l}{e}$), and handling effects ($\handle{e}{h}$). Abstractions come
in two forms: OCaml abstractions ($\lambda^o$) and C abstractions
($\lambda^c$). The handler consists of a return case ($\caseval{x}{e}$), zero
or more exception cases ($\caseexn{l}{x}{e}$) with label $l$, parameter $x$ and
body $e$, and zero or more effect cases ($\caseeff{l}{x}{k}{e}$) with label
$l$, parameter $x$, continuation $k$ and body $e$.

@@

We can take a C step (Figure~\ref{sem:cstep}) by taking an administrative
reduction step $\rightsquigarrow$. The administrative reductions are common to
both C and OCaml. The rules \textsc{Var}, \textsc{Arith1}, \textsc{Arith2},
\textsc{App1}, \textsc{App2} and \textsc{App3} are standard. \textsc{Arith3}
performs the arithmetic operation on the integers ($\llbracket n_1 \odot n_2
\rrbracket$). \textsc{Raise} pushes an function frame with exception value to
indicate that an exception is being raised. Similarly, \textsc{Perform} pushes
a function frame with an effect value with an empty continuation
$[[],(\{\caseval{x}{x}\},\emptyset)]$ with no captured frames and an empty
handler with an identity return case alone. We shall return to \textsc{Resume1}
and \textsc{Resume2} later.

@@

In the operational semantics, the continuations may be resumed more than once.
Captured continuations are copies of the original fibers and resuming the
continuation copies the fibers and leaves the continuation as it is. Since our
primary use case is concurrency, continuations will be resumed at-most once, and
copying fibers is unnecessary and inefficient. Instead, Multicore OCaml
optimises fibers for one-shot continuations. Fibers are allocated on the C heap
using |malloc| and are |free|d when the handled computation returns with a value
or an exception. Similar to Farvardin et al.~\cite{Farvardin20}, we use a
\emph{stack cache} of recently freed stacks in order to speed up allocation.

@@

Next, we have the variable-sized area for the OCaml frames. In order to keep
fibers small, this area is initially 16 words in length. When the stack pointer
|rsp| becomes less than the \emph{stack threshold} (maintained in the
|Caml_state| table), the stack is said to have overflown. On stack overflow, we
copy the whole fiber to a new area with double the size. In Multicore OCaml, we
introduce stack overflow checks into the function prologue of OCaml functions.
These stack overflows are rare and so the overflow checks will be correctly predicted
by the CPU branch predictor.

@@

As one of our primary goals of adding effect handlers to OCaml is to support
direct-style concurrency, we implement a full-fledged HTTP/1.1 web server using
effect handlers (|MC|). We use effect handlers to implement a user-level
threading library that spawns a thread per request. We use
\texttt{httpaf}~\cite{httpaf} for HTTP handling, and
\texttt{libev}~\cite{libev} for the eventloop. We compare our implementation
against an Lwt version (|lwt|) which also uses \texttt{httpaf} and
\texttt{libev}. Unlike using effect handlers, the Lwt version is written in
monadic style and does not have the notion of a thread per request. For
comparison, we include a Go 1.13 version (|go|) that uses the
\texttt{net/http}~\cite{nethttp} package. As both the OCaml versions are single
threaded, the Go benchmark is run with \texttt{GOMAXPROCS=1}.

@@

There are several strategies for implementing effect handlers. Eff~\cite{Eff},
Helium~\cite{Biernacki20}, Frank~\cite{Frank} and the Links server
backend~\cite{Hillerstrom20} use an interpreter similar to our operational
semantics to implement effect handlers. Effekt~\cite{Effekt}, Links JavaScript
backend~\cite{Hillerstrom20} and Koka~\cite{Leijen17} use type-directed
selective CPS translation. These language are equipped with an effect system,
which allows compiling pure code in direct-style and effectful code in CPS.
Leijen~\cite{Leijen14} implements effect handlers as a library in C using stack
copying. C allows pointers into the stack, so care is taken to ensure that
when continuations are resumed, the constituent frames are restored to the same
memory addresses as at the time of capture. Kiselyov et al.~\cite{Kiselyov18}
use an implementation of multi-prompt delimited continuations as an OCaml
library~\cite{Kiselyov12} to embed the Eff language in OCaml. Indeed, Forster et
al.~\cite{Forster19} showed that in an untyped setting, effect handlers,
monadic reflection and delimited control can macro-express each other.

@@

Our design for effect handlers in OCaml walks the tightrope of maintaining
compatibility (for profiling, debuggers and negligible overheads for existing
programs), while unlocking the full power of non-local control flow constructs.
Our evaluation shows that we have achieved our goal: we retain compatibility
with a surprisingly low performance overhead for sequential code that preserves
the spirit of ``fast exceptions'' that has always characterised OCaml
programming. We believe that the introduction of effect handlers into OCaml
implemented using lightweight fibers, along with a parallel
runtime~\cite{Sivaramakrishnan20}, will open OCaml to highly scalable
concurrent and task-parallel applications with minimal hit to sequential
performance.

>>>>

Effect handlers~\cite{Plotkin09} provide a modular foundation for user-defined
effects. The key idea is to separate the definition of the effectful operations
from their interpretations, which are given by \emph{handlers} of the effects.
For example,
%
\begin{lstlisting}
effect In_line : in_channel -> string
\end{lstlisting}
%
declares an \emph{effect} |In_line|, which is parameterised with an input
channel of type |in_channel|, which when \emph{performed} returns a |string|
value. A computation can perform the |In_line| effect without knowing how the
|In_line| effect is implemented. This computation may be enclosed by different
handlers that handle |In_line| differently. For example, |In_line| may be
implemented by performing a blocking read on the input channel or performing
the read asynchronously by offloading it to an event loop such as |libuv|,
without changing the computation. Thanks to the separation of effectful
operations from their implementation, effect handlers enable new approaches to
modular programming. Effect handlers are a generalization of exception
handlers, where, in addition to the effect being handled, the handler is
provided with the delimited continuation~\cite{Danvy90} of the perform site,
which may be resumed later. This enables non-local control-flow mechanisms such
as resumable exceptions, lightweight threads, coroutines, generators and
asynchronous I/O to be composably expressed.

One of the primary motivations to extend OCaml with effect handlers is to
natively support asynchronous I/O in order to express highly scalable
concurrent applications such as web servers in \emph{direct style} (as opposed
to using \emph{callbacks}). Many programming languages, including OCaml,
require non-local changes to source code in order to support asynchronous I/O,
often leading to a dichotomy between synchronous and asynchronous
code~\cite{colour}. For asynchronous I/O, OCaml developers typically use
libraries such as Lwt~\cite{lwt} and Async~\cite{async}, where asynchronous
functions are represented as monadic computations. In these libraries, while
asynchronous functions can call synchronous functions directly, the converse is
not true. In particular, any function that calls an asynchronous function will
also have to be marked as asynchronous. As a result, large parts of the
applications using these libraries end up being in monadic form. Languages such
as GHC Haskell and Go provide lightweight threads, which avoids the dichotomy
between synchronous and asynchronous code. However, these languages bake-in the
lightweight thread implementation into the runtime system. With effect
handlers, asynchronous I/O can be implemented directly in OCaml as a library
without imposing a monadic form on the users.

@@

OCaml currently does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. A standard way of implementing continuations is
to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Leijen14}. OCaml does not use a CPS IR, and changing
the compiler to utilise a CPS IR would be an enormous undertaking that would
affect the performance profile of existing OCaml applications due to the
increased memory allocations as the continuation closures get allocated on the
heap~\cite{Farvardin20}. Moreover, with CPS, an explicit stack is absent, and
hence, we would lose compatibility with tools that inspect the program stack.
Hence, we choose not to use CPS translation and represent the continuations as
call stacks.

@@

\begin{enumerate}[label=R\arabic*]
	\item \textbf{Backwards compatibility.} Extant OCaml programs \kc{Extant
		better than legacy?} do not break under OCaml extended with effect
		handlers. OCaml code that does not use effect handlers will pay negligible
		performance and memory cost.

@@

Our work is also timely. The WebAssembly~\cite{Hass17} community group is
considering effect handlers as one of the mechanisms for supporting
concurrency, asynchronous I/O and generators~\cite{WasmProposal}. Project
Loom~\cite{loom} is an OpenJDK project that adds virtual threads and delimited
continuations to Java. The Swift roadmap~\cite{swift} includes direct style
asynchronous programming and structured concurrency as milestones. We believe
that our design choices will inform similar choices to be made in other
industrial-strength languages.

@@

\begin{itemize}
	\item the design and implementation of effect handlers for OCaml. Our design
		retains OCaml's compatibility with program analysis tools that inspect the
		stack using DWARF unwind tables. We have validated our DWARF unwind tables
		with the assistance of an automated validator tool~\cite{Bastian19}.
	\item a formal operational semantics for the effect handler implementation in
		OCaml. Our formalism explicitly models the subtle interactions with the C
		stack, which is generally overlooked by other formal models, but which the
		implementations must handle.
	\item extensive evaluation which shows that our implementation has negligible
		impact on code that does not use effect handlers, and serves as an
		efficient foundation for scalable concurrent programming.
\end{itemize}

@@

OCaml does not use the C calling convention. In particular, there are no
callee-saved registers in OCaml. In the |x86-64| backend, the OCaml runtime
makes use of two C callee-saved registers for supporting OCaml execution. The
register |r15| holds the \emph{allocation pointer} into the minor heap used for
bump pointer allocation, and |r14| holds a reference to the |Caml_state|, a
table of global variables used by the runtime. This makes external calls
extremely fast in OCaml. If the external function does not allocate in the
OCaml heap, then it can be called directly and no bookkeeping is necessary. For
external functions which allocate in the OCaml heap, the cached allocation
pointer is saved to |Caml_state| before the external call and it is restored on
return. Similarly, callbacks into OCaml from C are also cheap: these involve
loading the arguments in the right registers and calling the OCaml function.
OCaml callbacks are relatively common as the garbage collector (GC), which is
implemented in C, executes OCaml finalisation functions as callbacks.

@@

The lack of callee-saved registers also makes exception handling fast. In the
absence of callee-saved registers, no registers need to be saved when entering
a |try| block. Similarly, no registers need to be restored when handling an
exception. Installing an exception handler simply pushes the program counter
(|pc|) of the handler and the current exception pointer (|exn_ptr| -- a field
in |Caml_state|) onto the stack. After this, the current exception pointer is
updated to be the current stack pointer (|rsp|). This creates a linked-list of
exception handler frames on the stack as shown in Figure~\ref{fig:stock_stack}.
Raising an exception simply sets |rsp| to |exn_ptr|, loads the saved |exn_ptr|,
and jumps to the |pc| of the handler.

@@

OCaml generates \emph{stack maps} in order to accurately identify roots on the
stack for assisting the GC. For every call point in the program, the OCaml
compiler emits the size of the frame and the set of all live registers in the
frame that point to the heap. During a GC, the OCaml stack is walked and the
roots are marked, skipping over the C frames.

@@

When the scheduler queue is empty, the |run_next| function performs the pending
reads. We abstract away the details of the event-based I/O using the |do_reads|
function (line 12). |do_reads| takes a list of pending reads and blocks until
at least one of the reads succeeds. It returns a pair of lists |compl| and
|todo|. |compl| contains the result strings from successful reads and
corresponding continuations. These continuations are arranged to be resumed
with the read result and pushed into the scheduler queue. |todo| contains the
channels on which input is still pending and their corresponding continuations.
|pending_reads| is updated to point to the |todo| list so that they may be
attempted later. Observe that all of the changes to add asynchrony are
localised to the |run| function, and the computation that performs these
effects can remain in direct style (as opposed to the monadic-style in
Lwt and Async).

@@

For handling the effects, our implementation extends OCaml's |match ... with|
syntax with effect patterns. The expression
\begin{lstlisting}
match e with
| None -> false | Some b -> b
| effect (E s) k1 -> e1 | effect (F f) k2 -> e2
\end{lstlisting}
\noindent is translated to the equivalent of
\begin{lstlisting}
match_with (fun () -> e)
{ retc = (function None -> false | Some b -> b);
  effc = (function
  | (E s) -> (fun k1 -> e1)
  | (F f) -> (fun k2 -> e2)
  | e -> (fun k -> match perform e with
          | v -> continue k v
          | exception e -> discontinue k e)); }
\end{lstlisting}
For the sake of exposition, we introduce a |('a,'b) handler| type. This handler
handles a |'a comp| that returns a |'a| value, and itself returns a |'b| value.
The handler has a return field |retc| of type |'a -> 'b|. The effect field
|effc| handles effects of type |'c eff| with |('c,'b) continuation| and returns
a value of type |'b|. The last case in |effc| \emph{reperforms} any unmatched
effect to the outer handler and returns the value and exceptions back to the
original performer. In the implementation, reperform is implemented as a
primitive to avoid executing code on the resumption path.

@@

\subsubsection{Syntax} Our expressions (Figure~\ref{sem:syntax}) consist of
integer constants ($n$), variables ($x$), abstraction ($\lam{x}{e}$),
application ($e~e$), arithmetic expressions ($e \odot e$) where $\odot$ ranges
over $\{+,-,*,/\}$, raising exceptions ($\throw{l}{e}$), performing effects
($\perform{l}{e}$), and handling effects ($\handle{e}{h}$). Abstractions come
in two forms: OCaml abstractions ($\lambda^o$) and C abstractions
($\lambda^c$). The handler consists of a return case ($\caseval{x}{e}$), zero
or more exception cases ($\caseexn{l}{x}{e}$) with label $l$, parameter $x$ and
body $e$, and zero or more effect cases ($\caseeff{l}{x}{k}{e}$) with label
$l$, parameter $x$, continuation $k$ and body $e$.

@@

We can take a C step (Figure~\ref{sem:cstep}) by taking an administrative
reduction step $\rightsquigarrow$. The administrative reductions are common to
both C and OCaml. The rules \textsc{Var}, \textsc{Arith1}, \textsc{Arith2},
\textsc{App1}, \textsc{App2} and \textsc{App3} are standard. \textsc{Arith3}
performs the arithmetic operation on the integers ($\llbracket n_1 \odot n_2
\rrbracket$). \textsc{Raise} pushes an function frame with exception value to
indicate that an exception is being raised. Similarly, \textsc{Perform} pushes
a function frame with an effect value with an empty continuation
$[[],(\{\caseval{x}{x}\},\emptyset)]$ with no captured frames and an empty
handler with an identity return case alone. We shall return to \textsc{Resume1}
and \textsc{Resume2} in the next subsection.

@@

In the operational semantics, the continuations may be resumed more than once.
Captured continuations are copies of the original fibers and resuming the
continuation copies the fibers and leaves the continuation as it is. Since our
primary use case is concurrency, continuations will be resumed at most once, and
copying fibers is unnecessary and inefficient. Instead, Multicore OCaml
optimises fibers for one-shot continuations. Fibers are allocated on the C heap
using |malloc| and are |free|d when the handled computation returns with a value
or an exception. Similar to Farvardin et al.~\cite{Farvardin20}, we use a
\emph{stack cache} of recently freed stacks in order to speed up allocation.

@@

Next, we have the variable-sized area for the OCaml frames. In order to keep
fibers small, this area is initially 16 words in length. When the stack pointer
|rsp| becomes less than the \emph{stack threshold} (maintained in the
|Caml_state| table), the stack is said to have overflowed. On stack overflow,
we copy the whole fiber to a new area with double the size. In Multicore OCaml,
we introduce stack overflow checks into the function prologue of OCaml
functions. These stack overflows are rare and so the overflow checks will be
correctly predicted by the CPU branch predictor.

@@

As one of our primary goals of adding effect handlers to OCaml is to support
direct style concurrency, we implement a full-fledged HTTP/1.1 web server using
effect handlers (|MC|). We use effect handlers to implement a user-level
threading library that spawns a thread per request. We use
\texttt{httpaf}~\cite{httpaf} for HTTP handling, and
\texttt{libev}~\cite{libev} for the eventloop. We compare our implementation
against an Lwt version (|lwt|) which also uses \texttt{httpaf} and
\texttt{libev}. Unlike using effect handlers, the Lwt version is written in
monadic style and does not have the notion of a thread per request. For
comparison, we include a Go 1.13 version (|go|) that uses the
\texttt{net/http}~\cite{nethttp} package. As both the OCaml versions are single
threaded, the Go benchmark is run with \texttt{GOMAXPROCS=1}.

@@

There are several strategies for implementing effect handlers. Eff~\cite{Eff},
Helium~\cite{Biernacki20}, Frank~\cite{Frank} and the Links server
backend~\cite{Hillerstrom20} use an interpreter similar to our operational
semantics to implement effect handlers. Effekt~\cite{Effekt}, Links JavaScript
backend~\cite{Hillerstrom20} and Koka~\cite{Leijen17} use type-directed
selective CPS translation. These language are equipped with an effect system,
which allows compiling pure code in direct style and effectful code in CPS.
Leijen~\cite{Leijen14} implements effect handlers as a library in C using stack
copying. C allows pointers into the stack, so care is taken to ensure that
when continuations are resumed, the constituent frames are restored to the same
memory addresses as at the time of capture. Kiselyov et al.~\cite{Kiselyov18}
use an implementation of multi-prompt delimited continuations as an OCaml
library~\cite{Kiselyov12} to embed the Eff language in OCaml. Indeed, Forster et
al.~\cite{Forster19} showed that in an untyped setting, effect handlers,
monadic reflection and delimited control can macro-express each other.

@@

Our design for effect handlers in OCaml walks the tightrope of maintaining
compatibility (for profiling, debuggers and negligible overheads for existing
programs), while unlocking the full power of non-local control flow constructs.
Our evaluation shows that we have achieved our goal: we retain compatibility
with a surprisingly low performance overhead for sequential code that preserves
the spirit of ``fast exceptions'' that has always characterised OCaml
programming. We believe that the introduction of effect handlers into OCaml
implemented using lightweight fibers, along with a parallel
runtime~\cite{Sivaramakrishnan20}, as has been demonstrated in our work, will
open OCaml to highly scalable concurrent and task-parallel applications with
minimal hit to sequential performance.
